Syntax and Semantics of Cedille

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: more small changes
- update authorship
- elaborate on bidirectional type inference
- introduce call-by-name normalization in type conversion

\author{Aaron Stump \\
  Computer Science \\
  The University of Iowa \\
\texttt{aaron-stump@uiowa.edu}}

@@

The type theory of Cedille is called the Calculus of Dependent Lambda
Eliminations (CDLE).  This document presents the version of CDLE as of
June 1, 2018.  We have made many changes from the first paper on CDLE~\cite{stump17},
mostly in the form of dropping constructs we discovered (to our
surprise) could be derived~\cite{stump18}. I have also omitted
\emph{lifting} -- a technique for large eliminations with lambda
encodings -- in this document's version of CDLE.  Some uses of lifting
can be simulated other ways within the system, though the limits of
this are still under investigation.  We also include a construct
$\delta$, for deriving a contradiction from a proof that
lambda-encoded true equals lambda-encoded false.  This also
compensates somewhat for the lack of lifting.

@@

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}.  For
brevity, we take these figures as implicitly specifying the syntax of
kinds $\kappa$, types $T$, and annotated terms $t$; these may use term
variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$. We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$).  The rules are
intended, with a few points of nondeterminism, to be read bottom-up
(in a standard way; cf.~\cite{peytonjones07}) as an algorithm for
computing a classifier from a context and an expression ($\tpsynth$)
or checking an expression against a classifier in context
($\tpcheck$).

The classification rules refer to an erasure function, defined in
Figure~\ref{fig:eraser}.  The type theory is \emph{extrinsic} (aka,
Curry-style), and hence we only consider erasures $|t|$ of terms when
testing for $\beta\eta$-equivalence.  This is done by the conversion
relation $T\cong T'$, whose central rules are given in
Figure~\ref{fig:conv}.  That figure omits the various congruence rules
needed to equate bigger expressions by equating subexpressions.  The
main ideas of conversion shown in the figure are to use
$\beta$-equivalence at the type level, and $\beta\eta$-equivalence of
erased terms at the term level.

@@

To Curry-style CC, CDLE adds: implicit products, introduced orginially
by Miquel~\cite{miquel01}; a primitive equality type $\{ t \simeq
t'\}$; and dependent intersection types $\abs{\iota}{x}{T}{T'}$,
introduced by Kopylov~\cite{kopylov03}.  Implicit products are used
for erased arguments to functions, found also in systems like Agda
(cf.~\cite{mishraLinger08}).  Dependent intersections are a rather
exotic construct allowing us to assign type $\abs{\iota}{x}{T'}{T}$ to
erased term $t$ when we can assign $T'$ to $t$, and also assign
$[t/x]T$ to $t$.  For an annotated introduction form, we write
$[t,t'$], where $t$ checks against type $T'$, $t'$ checks against
$[t/x]T$, and $t$ and $t'$ have identical (i.e., $\alpha$-equivalent)
erasures.  Dependent intersections thus enable a controlled form of
self-reference in the type.  Previous work showed how to use this to
derive induction for Church-encoded natural numbers~\cite{stump18}.
We will see below further uses of this construct.

The typing rules include conversion checks in a few places; e.g., as
standardly, when switching from checking to synthesizing mode.  Two
rules near the top of Figure~\ref{fig:tp} state that one may (nondeterministically)
$\beta$-reduce the type one is synthesizing or checking, before proceeding.
This allows reduction to head-normal form,
to match the form of type required by other rules.
Finally, we include the construct $\chi\ T\ \mhyph\ t$ to change the
synthesized or checked type $T'$ to $T$, if $T \cong T'$.  This may
be necessary to get the type into a specific form for purposes of rewriting
with the $\rho$ construct.

Finally, we have modified
the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t\}|$ erases to $|t|$.  We call this the \textbf{Kleene trick},
  because it goes back to Kleene's numeric realizability, which accepts any number $n$ as a realizer
  of a true equation.  Here, we accept any closed term $t$ as a realizer
  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term $t$.
\item The $\rho$ construct allows one to rewrite occurrences of $t_1$ to $t_2$ in the synthesized or
  checked type, where $t_1$ and $t_2$ are provably equal.  In the Cedille implementation, we rewrite
  all matching occurrences.  This may be compared to \verb|rewrite| in Agda, except that it may be
  applied anywhere, not just as part of pattern matching~\cite{agda}.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\phi\ t\ \mhyph\ t'\{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.  In the first version of CDLE, such an axiom was derivable from \emph{lifting}, a construct allowing
  simply typable terms to be lifted to the type level~\cite{stump17}.  We omit lifting in this new version of CDLE, because while
  sound, lifting as defined in that previous work is complicated and appears to be incomplete.  Developing a new
  form of lifting remains to future work.
\end{itemize}

@@

Finally, in line with ideas recently advocated by Dreyer, we
do not concern ourselves with syntactic
type preservation~\cite{dreyer18}, noting instead that by construction,
semantic types $\interp{T}_{\sigma,\rho}$ are preserved by $\beta\eta$-reduction:

>>>>

\author{Aaron Stump, Chris Jenkins \\
  Computer Science \\
  The University of Iowa \\
\texttt{aaron-stump@uiowa.edu}}

@@

The type theory of Cedille is called the Calculus of Dependent Lambda
Eliminations (CDLE).  This document presents the version of CDLE as of
December 1, 2020.  We have made many changes from the first paper on CDLE~\cite{stump17},
mostly in the form of dropping constructs we discovered (to our
surprise) could be derived~\cite{stump18}. I have also omitted
\emph{lifting} -- a technique for large eliminations with lambda
encodings -- in this document's version of CDLE.  Some uses of lifting
can be simulated other ways within the system, though the limits of
this are still under investigation.  We also include a construct
$\delta$, for deriving a contradiction from a proof that
lambda-encoded true equals lambda-encoded false.  This also
compensates somewhat for the lack of lifting.

@@

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}.  For
brevity, we take these figures as implicitly specifying the syntax of
kinds $\kappa$, types $T$, and annotated terms $t$; these may use term
variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
it is intended that such occurrences be read the same way (i.e., either all as
\(\tpcheck\) or all as \(\tpsynth\)).
We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$, in which case \(e\) must
be a term).
The rules are intended, with a few points of nondeterminism, to be read
bottom-up (in a standard way;
c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
an algorithm for synthesizing a classifier from a context and an expression
($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
($\tpcheck$) (\emph{type checking}).
Specifically, nondeterminism arises in the two rules allowing both synthesized
and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
constructors required by other rules may be revealed.
The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
is sufficient to reduce types to weak head normal form.

The classification rules refer to an erasure function, defined in
Figure~\ref{fig:eraser}.  The type theory is \emph{extrinsic} (aka,
Curry-style), and hence we only consider erasures $|t|$ of terms when
testing for $\beta\eta$-equivalence.  This applies also to the conversion
relation $T\cong T'$ on types and \(\kappa \cong \kappa'\) on kinds, whose central rules are given in
Figure~\ref{fig:conv}.  That figure omits the various congruence rules
needed to equate bigger expressions by equating subexpressions.  The
main ideas of conversion shown in the figure are to use
$\beta$-equivalence at the type level, and $\beta\eta$-equivalence of
erased terms at the term level.
Note in particular that for conversion we check full \(\beta\) and
\(\beta\eta\)-equivalence, not just equivalence with respect to call-by-name
normalization.

@@

To Curry-style CC, CDLE adds: implicit products, introduced orginially
by Miquel~\cite{miquel01}; a primitive equality type $\{ t \simeq
t'\}$; and dependent intersection types $\abs{\iota}{x}{T}{T'}$,
introduced by Kopylov~\cite{kopylov03}.  Implicit products are used
for erased arguments to functions, found also in systems like Agda
(cf.~\cite{mishraLinger08}).  Dependent intersections are a rather
exotic construct allowing us to assign type $\abs{\iota}{x}{T'}{T}$ to
erased term $t$ when we can assign $T'$ to $t$, and also assign
$[t/x]T$ to $t$.  For an annotated introduction form, we write
$[t,t'$], where $t$ checks against type $T'$, $t'$ checks against
$[t/x]T$, and $t$ and $t'$ have $\alpha\beta\eta$-equivalent
erasures.  Dependent intersections thus enable a controlled form of
self-reference in the type.  Previous work showed how to use this to
derive induction for Church-encoded natural numbers~\cite{stump18}.
We will see below further uses of this construct.

The typing rules include conversion checks in a few places; e.g., as
standardly, when switching from checking to synthesizing mode.  As mentioned
earlier, two rules near the top of Figure~\ref{fig:tp} state that one may
(nondeterministically) $\beta$-reduce the type one is synthesizing or checking,
before proceeding.
This allows reduction to weak head normal form, to match the form of type
required by other rules.
Finally, we include the construct $\chi\ T\ \mhyph\ t$ which allows a term whose
type can only be checked to be ascribed a user-provided type so that the whole
expression may occur where type synthesis is expected.

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to $|t'|$.  We call this the \textbf{Kleene trick},
  because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any closed term $t$ as a realizer
  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term $t$.
\item The $\rho$ construct allows one to rewrite occurrences of $t_1$ to $t_2$ in the synthesized or
  checked type, where $t_1$ and $t_2$ are provably equal.  In the Cedille implementation, we rewrite
  all matching occurrences.  This may be compared to \verb|rewrite| in Agda, except that it may be
  applied anywhere, not just as part of pattern matching~\cite{agda}.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\phi\ t\ \mhyph\ t'\{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.

  In the first version of CDLE, such an axiom was derivable from \emph{lifting}, a construct allowing
  simply typable terms to be lifted to the type level~\cite{stump17}.  We omit lifting in this new version of CDLE, because while
  sound, lifting as defined in that previous work is complicated and appears to be incomplete.  Developing a new
  form of lifting remains to future work.
\end{itemize}

@@

Finally, in line with ideas recently advocated by Dreyer, we
are less concerned with syntactic
type preservation as we are with \emph{semantic} type
preservation~\cite{dreyer18}.
Note that by construction, semantic types $\interp{T}_{\sigma,\rho}$ are
preserved by $\beta\eta$-reduction:
